---
main_topsize: 0.10 #percent coverage of the poster
main_bottomsize: 0.07
poster_height: "100cm"
poster_width: "70cm"
#ESSENTIALS
title: '**A holistic view of academic performance: Beyond Averages with MELSM and Spike-and-Slab**'
author:
  - name: '**Marwin Carmo**'
    affil: 1
    main: true
    #orcid: '0000-0002-1099-3857'
    email: mmcarmo@ucdavis.edu
  - name: Donald R. Williams
    affil: 1
    main: true
    email: 'drwwilliams@ucdavis.edu'
  - name: Philippe Rast
    affil: 1
    main: true
    email: 'prast@ucdavis.edu'
affiliation:
  - num: 1
    address: University of California, Daviss
# Main Colour Scheme
primary_colour: "#022851"
secondary_colour: "#00524C"
body_bgcol: "#ffffff"
#---POSTER BODY OPTIONS---#
title_textsize: "110px"
reference_textsize: "28px"
caption_textsize: "32px"
affiliation_textsize: "32px"
#Column
column_numbers: 2
#Main Section
main_fontfamily: "PT Sans"
main_textalign: "left"
main_textsize: "145px"
main_textalpha: "cf"
main_findings:
  - "**ivd**: a package for detecting intra individual differences"
main_textcol: "#ffffff"
logoleft_name: img/expanded_logo_white-gold.png
logoright_name: img/expanded_logo_white-gold.png
logocenter_name: img/expanded_logo_white.png
output: 
  posterdown::posterdown_betterport:
    highlight: NULL
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
    template: ../sono2021/poster/custom_template.html
#bibliography: references.bib
#link-citations: true
#csl: apa7.csl
nocite: '@*'
---

```{r setup, include=FALSE}
library(ivd) # local version
```


# Introduction

To get a more complete picture of academic achievement of a student or a school, it's important to look at both
average performance and variability within a cluster. We adapted the mixed-effects 
location scale model (MELSM) using the Spike and  Slab regularization technique 
to shrink random effects to their fixed effect. 
This allows us to identify clusters with unusually (in)consistent academic achievement. 

An indicator variable ($\delta$) is included in the prior to allow switching between the spike and slab  throughout the MCMC sampling process. For each iteration, this specification allows each individual to have their own person-specific estimate or the fixed effect average. 

The methods have been implemented in the R package `ivd`:

```{r eval=FALSE, include=TRUE}
devtools::install_github("consistentlybetter/ivd")
```


# Illustrative example

The data comes from The Basic Education Evaluation System (Saeb) conducted by Brazil’s National Institute for Educational Studies and Research (Inep). It is also available as the `saeb` dataset in the `ivd` package. The outcome variable is `math_proficiency` at the end of grade 12. Both, location and scale are modeled as a function of student and school SES. 

For student $i$ in school $j$, we model the location as,

$$
\text{Math}_{ij} = \beta_0+\beta_1\text{SES}_{ij}^{\text{std}}+\beta_2\text{SES}_{ij}^{\text{scl}}+\beta_3\text{SES}_{ij}^{\text{std}}\times\text{SES}_{ij}^{\text{std}} + u_j + e_{ij}
$$
$$
e_{ij} \sim \mathcal{N} (0, \sigma_{\epsilon_{ij}}^2)
$$
And the scale as,

$$
\sigma_{\epsilon_{ij}}^2 = \eta_0+\eta_1\text{SES}_{ij}^{\text{std}}+\eta_2\text{SES}_{ij}^{\text{scl}}+\eta_3\text{SES}_{ij}^{\text{std}}\times\text{SES}_{ij}^{\text{std}} + t_j
$$
#
