---
main_topsize: 0.10 #percent coverage of the poster
main_bottomsize: 0.07
poster_height: "100cm"
poster_width: "75cm"
#ESSENTIALS
title: '**A holistic view of academic performance: Beyond Averages with MELSM and Spike-and-Slab**'
author:
  - name: '**Marwin Carmo**'
    affil: 1
    main: true
    #orcid: '0000-0002-1099-3857'
    email: mmcarmo@ucdavis.edu
  - name: Donald R. Williams
    affil: 1
    main: true
    email: 'drwwilliams@ucdavis.edu'
  - name: Philippe Rast
    affil: 1
    main: true
    email: 'prast@ucdavis.edu'
affiliation:
  - num: 1
    address: University of California, Daviss
# Main Colour Scheme
primary_colour: "#022851"
secondary_colour: "#00524C"
body_bgcol: "#ffffff"
#---POSTER BODY OPTIONS---#
title_textsize: "110px"
reference_textsize: "28px"
caption_textsize: "32px"
affiliation_textsize: "32px"
#Column
column_numbers: 2
#Main Section
main_fontfamily: "PT Sans"
main_textalign: "left"
main_textsize: "145px"
main_textalpha: "cf"
main_findings:
  - "**ivd**: an R package for individual variance detection"
main_textcol: "#ffffff"
logoleft_name: img/imps_logo.png
logoright_name: img/expanded_logo_white-gold.png
logocenter_name: img/qrcode.svg
output: 
  posterdown::posterdown_betterport:
    highlight: NULL
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
    template: ../sono2021/poster/custom_template.html
bibliography: references.bib
link-citations: true
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
library(ivd) # local version
library(ggplot2)
out <- readRDS("out/out.rds")
s_out <- summary(out)
source("plotivd.R")
```


# Introduction

To get a more complete picture of academic achievement of a student or a school, it’s important to look at both average performance and variability within a cluster. We adapted the mixed-effects location scale model
(MELSM) using the Spike and Slab regularization technique to shrink the *scale* random effects to their fixed effect. This allows us to identify clusters with unusually (in)consistent academic achievement.

### Model formulation

The random effects from the scale and location can be modeled as, $\textbf{u}\sim\mathcal{N}(\textbf{0}, \boldsymbol{\Sigma})$. The covariance matrix of the random effects can be decomposed into $\boldsymbol\Sigma = \boldsymbol{\tau}\boldsymbol{\Omega\tau}'$ to specify independent priors for each element of $\boldsymbol{\tau}$ (random-effects standard deviations) and $\boldsymbol{\Omega}$ (correlation matrix among all random effects).

$\boldsymbol\Omega$ can be factorized via the Cholesky $\textbf{L}$ of $\boldsymbol\Omega = \textbf{L}'\textbf{L}$. Then, $\textbf{u}$ can be recovered multiplying $\textbf{L}$ by the random effect standard deviations, $\boldsymbol{\tau}$, and scaling it with a standard normally distributed $\boldsymbol{z}$.

An indicator variable ($\delta$) is included in the prior for the scale random effect to allow switching between the spike and slab throughout the MCMC sampling process. For each school $j$, we obtain a probabilistic measure on whether the scale random effect is to be included or not: 

$$\textbf{u}_j = \boldsymbol{\tau}\textbf{L}\color{red}{\boldsymbol{\delta}}_j\textbf{z}_j$$
For a random intercept only model,

\begin{equation}
\label{eq:mm_delta}
\alpha_{0j} = \begin{cases}
\eta_{0}, & \text{if }\delta_j = 0 , \\
\eta_{0} + u_{0j}, & \text{if }\delta_j = 1
\end{cases}.
\end{equation}

The posterior inclusion probabilities (PIP) can then be computed as

\begin{align}
Pr(\alpha_{0j} = \eta_{0} | \textbf{Y}) = 1 - \frac{1}{S} \sum_{s = 1}^S \delta_{js},
\end{align}

where $S = \{1,...,s\}$ denotes the posterior samples and $\eta_{0}$ is the average within-person variability.


The methods have been implemented in the R package [`ivd`](https://github.com/consistentlyBetter/ivd):

```{r eval=FALSE, include=TRUE}
devtools::install_github("consistentlybetter/ivd")
```



# Illustrative example

The data comes from The Basic Education Evaluation System (Saeb) conducted by Brazil's National Institute for Educational Studies and Research (Inep). It is also available as the `saeb` dataset in the `ivd` package. The outcome variable is `math_proficiency` at the end of grade 12. Location and scale are modeled as a function of student and school SES. 


```{r plots, include=FALSE}
p2 <- plot.ivd(out, "funnel")
p3 <- plot.ivd(out, "outcome")
```


```{r funnel-plot, echo=FALSE,dpi=300,out.width="80%", fig.align='center', fig.cap="Funnel plot of posterior inclusion probabilities and school's standard deviations. Higher PIP values express more evidence for the random effect of a given school to be included in the model."}
p2 +
  theme_minimal()
```




```{r y-plot, echo=FALSE, dpi=300,out.width="80%",fig.align='center', fig.cap="Posterior inclusion probabilities plotted against school's average math achievement in standardized units. Point size represents within-school standard deviation."}
p3 +
  theme_minimal() +
  labs(x = "Math achievement (z-standardized)")
```
# Discussion

- Individual variance detection analysis suggests that some schools have students consistently performing above the average, while others have an average mean score but with greater-than-average variability. Inconsistent performance may reflect unaccounted factors influencing learning.

- The IVD method could also be used to assess academic performance throughout the year and uncover information relevant to providing more specific school or student support.

## References