---
title: "A network approach of Insomnia and Dysfunctional Beliefs and Attitudes About Sleep"
subtitle: "VII Congresso Clínica Psiquiátrica 2022"
author: 
  - "Marwin M I B Carmo"
  - "Renatha El Rafihi Ferreira"
date: "<img src ='img/fapesp-ipq.png' width = '30%'>"
institute: "Department of Psychiatry, Faculty of Medicine, University of São Paulo, Brazil"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "css/custom.css"]
    nature:
      #titleSlideClass: ["right", "top", "custom.css"]
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---
layout: true
  
<div class="my-footer"><span>

<a href="https://marwincarmo.github.io">Marwin M I B Carmo</a> (<a href="mailto:marwin@usp.br">marwin@usp.br</a>) -- A network approach of Insomnia and Dysfunctional Beliefs and Attitudes About Sleep.
</span></div>

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	fig.height = 4,
	fig.retina = 3,
	fig.width = 9,
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	hiline = TRUE,
	out.width = "100%"
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(xaringanExtra)
library(dplyr)
library(ggplot2)
library(bootnet)
library(kableExtra)
library(patchwork)

style_duo_accent(
  title_slide_background_color ="#FFFFFF",
  title_slide_text_color = "#024F84",
  primary_color = "#024F84",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF",
  text_color = "#515151",
  text_font_size = "1.4rem",
  table_row_even_background_color = "#D6EAF8"# "#a1c7cf",
)
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(
  image_url = c("img/fapesp-ipq.png"),
  width = "365px",
  height = "79px",
  position = css_position(top = "1em", right = "1em"),
  exclude_class = c("my-title", "inverse", "hide_logo")
)
```

```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble()
```


```{r database , include=FALSE}
raw_data <- readr::read_csv("data_dbas.csv")

```

```{r table1-data}
descriptive_data <- raw_data %>% 
  dplyr::select(sexo, etnia, escolaridade, medicacao_semana, regiao, group) %>% 
  tidyr::pivot_longer(cols = c(sexo, etnia,sexo, etnia, escolaridade, medicacao_semana, regiao, group),
                      names_to = 'variable', values_to = 'value',
                      values_transform = list(value=as.character)) %>% 
  dplyr::with_groups(c(variable, value), count) %>% 
  dplyr::with_groups(c(variable), mutate, p = n/sum(n)) %>% 
  dplyr::mutate(vars = dplyr::case_when(
    variable == "escolaridade" & value %in% c(1:7) ~ "Less than 12th grade",
    variable == "escolaridade" & value == 8 ~ "Some college",
    variable == "escolaridade" & value == 9 ~ "College degree or higher",
    variable == "etnia" & value == 1 ~ "White",
    variable == "etnia" & value %in% c(2,3) ~ "Black",
    variable == "etnia" & value == 4 ~ "Asian",
    variable == "etnia" & !(value %in% c(1:4)) ~ "Other/Not informed",
    variable == "medicacao_semana" & value == 0 ~ "Don't use",
    variable == "medicacao_semana" & value %in% c(1:5) ~ "1-5 days a week",
    variable == "medicacao_semana" & value %in% c(6,7) ~ "6-7 days a week",
    variable == "regiao" & value == "Região Centro-Oeste" ~ "Central-West",
    variable == "regiao" & value == "Região Nordeste" ~ "Northeast",
    variable == "regiao" & value == "Região Norte" ~ "North",
    variable == "regiao" & value == "Região Sudeste" ~ "Southeast",
    variable == "regiao" & value == "Região Sul" ~ "South",
    variable == "sexo" & value == "F" ~ "Female",
    variable == "sexo" & value == "M" ~ "Male",
    TRUE ~ value
  )) %>% 
  dplyr::with_groups(c(variable, vars), summarise, n = sum(n), p = sum(p)) %>% 
  #dplyr::mutate(total = paste0(n, " (", signif(p*100,3), ")")) %>%
  dplyr::mutate(variable = factor(variable, 
                                  levels = c("sexo", "etnia", "escolaridade", "regiao", "medicacao_semana", "group"))) %>% 
  dplyr::arrange(variable) #%>%
  #dplyr::select(-c(variable, n, p))


continuous_data <- raw_data %>% 
  dplyr::select(idade, renda) %>% 
  dplyr::summarise(mean_age = mean(idade, na.rm = TRUE),
                   sd_age = sd(idade, na.rm = TRUE),
                   income = median(renda, na.rm = TRUE))

```


```{r bases}
# By item
base_items <- raw_data |> 
  dplyr::filter(redcap_event_name == "elegibilidade_arm_1") |> 
  dplyr::select(igi_escore, dbas_1:dbas_16) |> 
  dplyr::rename(ISI = igi_escore,
                Q1 =dbas_1, Q2 =dbas_2,
                Q3 =dbas_3, Q4 =dbas_4,
                Q5 =dbas_5, Q6 =dbas_6,
                Q7 =dbas_7, Q8 =dbas_8,
                Q9 =dbas_9, Q10 =dbas_10, 
                Q11=dbas_11, Q12 = dbas_12,
                Q13=dbas_13, Q14 = dbas_14,
                Q15=dbas_15, Q16 = dbas_16)
items <- c("Insomnia Severity Index",
           "I need 8 hours of sleep to feel refreshed and function well during the day",
           "When I don't get proper amount of sleep on a given night, \n I need to catch up on the next day by napping or on the next night by sleeping longer",
           "I am concerned that chronic insomnia may have serious consequences on my physical health",
           "I am worried that I may lose control over my abilities to sleep",
           "After a poor night’s sleep, I know that it will interfere with \n my daily activities on the next day",
           "In order to be alert and function well during the day, \n I believe I would be better off taking a sleeping pill rather than having a poor night’s sleep",
           "When I feel irritable, depressed, or anxious during the day,\n it is mostly because I did not sleep well the night before",
           "When I sleep poorly on one night, I know it will disturb my sleep schedule for the whole week",
           "Without an adequate night’s sleep, I can hardly function the next day",
           "I can’t ever predict whether I’ll have a good or poor night’s sleep",
           "I have little ability to manage the negative consequences of disturbed sleep",
           "When I feel tired, have no energy, or just seem not to function well during the day,\n it is generally because I did not sleep well the night before",
           "I believe insomnia is essentially the result of a chemical imbalance",
           "I feel insomnia is ruining my ability to enjoy life and prevents me from doing what I want",
           "Medication is probably the only solution to sleeplessness",
           "I avoid or cancel obligations (social, family) after a poor night’s sleep")
           

factors <- c("Insomnia Severity Index", rep("Expectations", 2), rep("Worry", 2), "Consequences",
             "Medication", "Consequences", "Worry", "Consequences", rep("Worry", 2), "Consequences",
             "Medication", "Worry", "Medication", "Consequences")

Network_items <- estimateNetwork(base_items, default = "EBICglasso", weighted = TRUE)

# By factors

base_factors <- raw_data |> 
  dplyr::filter(redcap_event_name == "elegibilidade_arm_1") |> 
  dplyr::select(igi_escore, dbas_1:dbas_16) |> 
  dplyr::mutate(cons = c(dbas_5 + dbas_7 + dbas_9 + dbas_12 + dbas_16),
               worry = c(dbas_3 + dbas_4 + dbas_8 + dbas_10 + dbas_11 + dbas_14),
               exp = c(dbas_1 + dbas_2),
               med = (dbas_6 + dbas_13 + dbas_15)
               ) |> 
  dplyr::select(-c(dplyr::starts_with("dbas_"))) |> 
  dplyr::rename(isi = igi_escore)

Network_factors <- estimateNetwork(base_factors, default = "EBICglasso", weighted = TRUE)
```

---
# Introduction
.v-center[
Negatively toned cognitive activity triggers arousal and distress, channeling attention and monitoring to sleep, and creating distorted perceptions ([Harvey, 2002](https://doi.org/fwxq35)).

Challenging unhelpful beliefs about sleep is a crucial element of cognitive-behavioral therapy for insomnia.

Understanding how this network of beliefs connects to insomnia severity may provide leads to help targeted interventions.
]
---

# Introduction
.pull-left[

A **network** structure is represented by **nodes** (variables), connected by **edges** (strength of associations) ([Burger et al., 2022](https://doi.org/10.1037/met0000471)).

Measures of centrality ([Borsboom et al., 2021](https://doi.org/10.1038/s43586-021-00055-w)):

- **Node strenght**: sums the absolute edge weights of edges per node;  
- **Closeness**: quantifies the distance between the node and all other nodes;  
- **Betweenness**: quantifies how often a node lies on the shortest path connecting any two other nodes.

]

.pull-right[

```{r, fig.height=5, fig.width=6, out.width="100%"}
mat <- matrix(rep(0,5*5), nrow=5)
mat[lower.tri(mat)] <-  c(.2,-.8,0,.6,
                         -.3,.3,.9,
                         .2,.4,
                         0)
mat <- t(mat)
mat[lower.tri(mat)] <-  c(.2,-.8,0,.6,
                          -.3,.3,.9,
                          .2,.4,
                          0)
qgraph::qgraph(mat, layout = "spring", edge.labels = TRUE, 
               labels = LETTERS[1:5], fade = TRUE)

```


]

---
# Objective
.v-center[

To use a network approach to explore interactions between dysfunctional beliefs and attitudes about sleep and insomnia severity.

To identify specific nodes within dysfunctional beliefs and attitudes about sleep that may play a pivotal role in maintaining insomnia.
]
---

# Method
.v-center[
**Participants**: 1166 adults, aged 18 to 59 years, both with and without insomnia symptoms.

**Instruments**:

- Insomnia Severity Scale (ISI);

- Dysfunctional Beliefs and Attitudes about Sleep Scale (DBAS-16). 
]
---
# Method
.v-center[
We fitted Gaussian graphical models using graphical LASSO regularization in combination with Extended Bayesian Information Criterion (EBIC) model selection, with a tuning hyperparameter set to 0.5.

- A network describing the associations among overall insomnia severity and the four **factors** of DBAS-16

- A network describing the associations among overall insomnia severity and each of the individual **items** of DBAS-16.

The stability of node strength was estimated using a case dropping bootstrap procedure (1000 iterations). Node centrality was assessed by strength, closeness, and betweenness.
]
---

# Sample Characteristics

.pull-left[
- **Mean Age (SD)**: `r round(continuous_data["mean_age"], 2)` (`r round(continuous_data["sd_age"], 2)`)

- **Gender (Female)**: `r subset(descriptive_data, n, subset=(vars=="Female"))` (`r round(subset(descriptive_data, p, subset=(vars=="Female"))*100, 2)`%)

- **Insomnia symptoms**: `r subset(descriptive_data, n, subset=(vars=="Insomnia"))` (`r round(subset(descriptive_data, p, subset=(vars=="Insomnia"))*100, 2)`%)

- **Race/ethnicity**:
  - Asian: `r subset(descriptive_data, n, subset=(vars=="Asian"))` (`r round(subset(descriptive_data, p, subset=(vars=="Asian"))*100, 2)`%)
  - Black: `r subset(descriptive_data, n, subset=(vars=="Black"))` (`r round(subset(descriptive_data, p, subset=(vars=="Black"))*100, 2)`%)
  - White: `r subset(descriptive_data, n, subset=(vars=="White"))` (`r round(subset(descriptive_data, p, subset=(vars=="White"))*100, 2)`%)
  - Other/Not informed: `r subset(descriptive_data, n, subset=(vars=="Other/Not informed"))` (`r round(subset(descriptive_data, p, subset=(vars=="Other/Not informed"))*100, 2)`%)
  
]

.pull-right[

- **Education**:
  - College degree or higher: `r subset(descriptive_data, n, subset=(vars=="College degree or higher"))` (`r round(subset(descriptive_data, p, subset=(vars=="College degree or higher"))*100, 2)`%)
  - Less than 12th grade: `r subset(descriptive_data, n, subset=(vars=="Less than 12th grade"))` (`r round(subset(descriptive_data, p, subset=(vars=="Less than 12th grade"))*100, 2)`%)
  - Some college: `r subset(descriptive_data, n, subset=(vars=="Some college"))` (`r round(subset(descriptive_data, p, subset=(vars=="Some college"))*100, 2)`%)

- **Region of origin**:
  - Central-West: `r subset(descriptive_data, n, subset=(vars=="Central-West"))` (`r round(subset(descriptive_data, p, subset=(vars=="Central-West"))*100, 2)`%)
  - North: `r subset(descriptive_data, n, subset=(vars=="North"))` (`r round(subset(descriptive_data, p, subset=(vars=="North"))*100, 2)`%)
  - Northeast: `r subset(descriptive_data, n, subset=(vars=="Northeast"))` (`r round(subset(descriptive_data, p, subset=(vars=="Northeast"))*100, 2)`%)
  - South: `r subset(descriptive_data, n, subset=(vars=="South"))` (`r round(subset(descriptive_data, p, subset=(vars=="South"))*100, 2)`%)
  - Southeast: `r subset(descriptive_data, n, subset=(vars=="Southeast"))` (`r round(subset(descriptive_data, p, subset=(vars=="Southeast"))*100, 2)`%)
]
---
# Results

```{r factors-plot, fig.cap = "isi = Insomnia Severity Index, worry = Worry about sleep, exp = Sleep expectations, cons = Consequences of insomnia, med = Medication.", out.width="80%"}
plot(Network_factors, layout = "spring",
     theme = "colorblind",color="seashell2",
     nodeNames = colnames(base_factors), 
     legend.mode="style2", label.cex = 1.2,
     vsize = 12, esize = 15, details = F,legend = FALSE)

```

---

# Results

```{r centrality-factors-plot}
source("centrality_plots.R")
plot(cntPlot1)
```

---
# Results

```{r edge-estimates-factors, message=FALSE, warning=FALSE, include=FALSE}

# Accuracy and  of Edge-Estimates
boot1 <- bootnet(Network_factors, nBoots = 10)

p1 <- plot(boot1, labels = FALSE,
     order = "sample")

# Stability

boot2 <- bootnet(Network_factors, nBoots = 10, type = "case", statistics = c("strength", "closeness", "betweenness"))

p2 <- plot(boot2, statistics= c("strength","closeness","betweenness")) 

```

```{r plot-edge-estimates-factors}
p1 + p2
```

---

# Results

```{r}
plot(Network_items, layout = "spring",
     label.cex=.70, label.scale=F, theme = "colorblind",
     nodeNames = items, groups = factors,
     legend.mode="style2", legend.cex=.22,
     vsize = 6, esize = 15, details = F, 
     layoutOffset = c(-.2,0))
```
---
# Results

```{r centrality-items-plot}

plot(cntPlot2)
```

---

# Results

```{r edge-estimates-items, message=FALSE, warning=FALSE, include=FALSE}

# Accuracy and  of Edge-Estimates
boot3 <- bootnet(Network_items, nBoots = 10)

p3 <- plot(boot3, labels = FALSE,
     order = "sample")

# Stability

boot4 <- bootnet(Network_items, nBoots = 10, type = "case", statistics = c("strength", "closeness", "betweenness"))

p4 <- plot(boot4, statistics= c("strength","closeness","betweenness")) 

```

```{r plot-edge-estimates-items}
p3 + p4
```

---
# Conclusion

- We used Network analysis as an exploratory technique to identifies patterns of conditional independence that may help interpreting empirical phenomena.

- *Sleep expectations* is negatively associated with overall insomnia severity score.

- **Worry about sleep** is the most central item in terms of strength, indicating its importance as possible target for cognitive-behavioral therapy for insomnia.

- *Belief that a poor night of sleep would ruin functioning on the next day* and *feelings that insomnia is ruining ability to enjoy life*
---
# References

```{r eval=FALSE, include=FALSE}
doi <- c('https://doi.org/10.1037/met0000471', 'https://doi.org/10.1038/s43586-021-00055-w')
#cat(unlist(rcrossref::cr_cn(dois = doi, format = "text", style = 'apa')))
refs <- rcrossref::cr_cn(dois = doi, format = "text", style = 'apa')
```